The Route 508 Streetcar and Congestion on Toronto's Shoreline, 2011 - 2020

### 2024-04-18 paper/paper.qmd
completed discussion and model results, updated references and included literature

---
title: "The Route 508 Streetcar and Congestion on Toronto's Shoreline, 2011 - 2019"
subtitle: "Examining Traffic Patterns Before and After Public Transit Access was Reinstated"
author: 
  - Maria Mangru
thanks: "Code and data are available at: https://github.com/MariaMangru/Traffic-Congestion-on-Toronto-s-Shoreline."
date: today
date-format: long
abstract: "This study investigates the impact of access to the Route 508 Lake Shore streetcar in Toronto on reducing urban car congestion from 2011 to 2020. By analyzing traffic volume data from the City of Toronto's Transportation Services Division and employing negative binomial regression models, we observed a notable shift in travel behavior, highlighted by a significant decrease in the correlation between bus traffic and car congestion. Specifically, the coefficient for daily bus traffic reduced from 0.261 to 0.096 after access was reinstated, indicating a reduction in car traffic as more public transit options became available. This highlights the effectiveness of enhancing public transit infrastructure not only in easing car congestion but also in contributing to broader environmental and economic benefits. By improving public transit, cities can potentially achieve sustainable development goals by providing equitable access to essential services and reducing environmental impacts, thereby improving quality of life for all urban residents."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r, echo=FALSE, message=FALSE}
#| include: false
#| warning: false
#| message: false

library(dplyr)
library(ggplot2)
library(tidyr)
library(readr)
library(here)
library(knitr)
library(kableExtra)
library(stringr)
library(broom)
library(MASS) 
library(car)
```

@@

The primary aim of this research is to evaluate the effect of the 508 Lake Shore streetcar enhancements on daily traffic patterns, focusing on how these changes influence the interaction between car traffic and other transportation modes such as buses, bicycles, and pedestrians. By employing traffic volume data collected by the City of Toronto's Transportation Services Division, we utilize negative binomial regression models to quantify the shifts in traffic patterns from 2011 to 2020. The estimand of this study is the change in the correlation between daily bus traffic and daily car traffic following the transit improvements.

@@

This research relies on traffic volume provided by the City of Toronto's Transportation Services Division, available through the Toronto OpenData portal. The dataset, titled "Traffic Volumes at Intersections for All Modes," is an open-access resource under the Open Government License - Toronto, ensuring broad usability and transparency. The data files for the years 2010-2019 were utilized, however, the data files are uploaded daily with data ranging from 1980 to present. The traffic volume data are cross-referenced with the City of Toronto's Street Centreline, Intersection File, and Street Traffic Signal datasets.

@@

The analysis in this paper was conducted using the R programming language @citeR, which facilitated the statistical computations and data visualization. Further, a suite of R packages, including several from the tidyverse collection were utilized. The `dplyr` package was used the transformation and summarization of traffic data. For importing and processing string data, `readr` and `stringr` were utilized. The handling of date and time variables was done using `lubridate` and data visualization was achieved with `ggplot2`, allowing us to explore and reveal traffic patterns visually. Additionally, `tidyr` assisted in reshaping the data to ensure it was structured appropriately for analysis. File path management was handled using the `here` package, which was essential for ensuring reproducibility across different computing environments.

The dataset was refined to aggregate daily totals for each mode of transportation, with the data aligned and merged based on date, location, and traffic mode specifics.

@@

In addition to a temporal trend variables, specific variables were chosen for this analysis, with an emphasis on the dynamics of car traffic in relation to bus, pedestrian, and bicycle traffic volumes:

-   Daily Car Traffic (`daily_cars`): Total volume of car traffic recorded at selected intersections along the 508 Route.
-   Daily Bus Traffic (`daily_bus`), Pedestrian Traffic (`daily_peds`) and Bicycle Traffic (`daily_bike`): Bus, pedestrian and bike traffic volumes used to indicate the presence and frequency of non personal car usage.
-   Time Trend (time_trend): A constructed variable to analyze traffic pattern changes over time.

```{r summary-stats-table, echo=FALSE, message=FALSE, warning=FALSE}
# Load the cleaned data
cleaned_data <- read_csv(here("data", "analysis_data", "cleaned-data.csv"))

# Generate summary statistics for traffic data
summary_stats <- summarise(cleaned_data,
  Mean_Cars = mean(daily_cars, na.rm = TRUE),
  Median_Cars = median(daily_cars, na.rm = TRUE),
  Mean_Buses = mean(daily_bus, na.rm = TRUE),
  Median_Buses = median(daily_bus, na.rm = TRUE),
  Mean_Pedestrians = mean(daily_peds, na.rm = TRUE),
  Median_Pedestrians = median(daily_peds, na.rm = TRUE),
  Mean_Bicycles = mean(daily_bike, na.rm = TRUE),
  Median_Bicycles = median(daily_bike, na.rm = TRUE)
)

@@

The summary table above provides a quantitative comparison across transportation modes by condensing the traffic data for each mode into average and median statistics. The data reveals that car traffic surpasses other modes in volume, with the mean and median values for daily car traffic being the highest.

```{r line-graph-cars, fig.cap="Daily Car Traffic Over Time. This graph displays the trend in number of cars within the area of interest between December 2011 to December 2014 and November 2018 to February 2020.",fig.width=8, fig.height=4, echo=FALSE, message=FALSE, warning=FALSE}

# Visualizing daily car traffic over time
line_graph_car <- cleaned_data %>%
  ggplot(aes(x = count_date, y = daily_cars)) +
  geom_smooth(se = FALSE) +  
  labs(title = "Daily Cars Traffic Over Time",
       x = "Date",
       y = "Total Cars Traffic") +
  theme_minimal()

# Display car traffic plot
line_graph_car

@@

The daily variation in car traffic from 2011 to 2020 is shown in Figure 1. The traffic volume is depicted on a smooth line graph that first increases and peaks about 2016, after which it sharply declines. The fall that follows may point to the benefits of the Route 508 Lake Shore streetcar improvements on easing traffic congestion. The peak denotes a period of increased traffic.

```{r line-graph-bus, fig.cap="Daily Bus Traffic Over Time. This graph displays the trend in number of buses within the area of interest between December 2011 to December 2014 and November 2018 to February 2020.", fig.width=8, fig.height=4, echo=FALSE, message=FALSE, warning=FALSE}

# Visualizing daily bus traffic over time
line_graph_buses <- cleaned_data %>%
  ggplot(aes(x = count_date, y = daily_bus)) +
  geom_smooth(se = FALSE) +
  labs(title = "Daily Bus Traffic Over Time",
       x = "Date",
       y = "Total Bus Traffic") +
  theme_minimal()

# Display bus traffic plot
line_graph_buses

Figure 2 depicts the daily patterns in bus traffic, which follow a similar pattern to that of car traffic. It is a smooth line graph that rises to a high around in 2016, then declines. This pattern might be an indication of how bus utilization has changed, which could be related to the streetcar route's service modifications and other improvements to public transportation.

```{r line-graph-pedestrian, fig.cap="Daily Pedestrian Traffic Over Time. This graph displays the trend in number of pedestrian within the area of interest between December 2011 to December 2014 and November 2018 to February 2020.", fig.width=8, fig.height=4, echo=FALSE, message=FALSE, warning=FALSE}

# Visualizing daily pedestrian traffic over time
line_graph_pedestrians <- cleaned_data %>%
  ggplot(aes(x = count_date, y = daily_peds)) +
  geom_smooth(se = FALSE) +
  labs(title = "Daily Pedestrian Traffic Over Time",
       x = "Date",
       y = "Total Pedestrian Traffic") +
  theme_minimal()

# Display pedestrian traffic plot
line_graph_pedestrians

Variations in pedestrian traffic over is shown in Figure 3's line graph. The middle of the time period has a noteworthy decrease, which is followed by a rise and another decline. This could be attributed to a number of things, such as changes in walkability, urban development, and perhaps the availability of better public transportation alternatives that encourage walking as a mode of transportation.

```{r line-graph-bike, fig.cap="Daily Bike Traffic Over Time. This graph displays the trend in number of bikes within the area of interest between December 2011 to December 2014 and November 2018 to February 2020.", fig.width=8, fig.height=4, echo=FALSE, message=FALSE, warning=FALSE}

# Visualizing daily bike traffic over time
line_graph_cyclists <- cleaned_data %>%
  ggplot(aes(x = count_date, y = daily_bike)) +
  geom_smooth(se = FALSE) +
  labs(title = "Daily Cyclist Traffic Over Time",
       x = "Date",
       y = "Total Cyclist Traffic") +
  theme_minimal()

# Display cyclist traffic plot
line_graph_cyclists

The pattern of bicycle traffic over time is depicted in Figure 4, where a notable rise and fall are shown on the line graph, indicating huge variations. The variations may be attributed to changing public opinions on riding, the implementation of bike infrastructure, or seasonal variations in weather patterns, which may be exacerbated by changes made to the transit system.

```{r line-graph-time-trend, fig.cap="Daily Car Traffic Over Time. This graph displays the general trend in number of cars from the beginning of data collection period.", fig.width=8, fig.height=4, echo=FALSE, message=FALSE, warning=FALSE}

# Visualizing car traffic with respect to the time trend
time_trend_graph_car <- cleaned_data %>%
  ggplot(aes(x = time_trend, y = daily_cars)) +
  geom_point(aes(color = daily_cars), alpha = 0.5) + 
  geom_smooth(method = "lm", se = FALSE, color = "blue") +  
  scale_color_gradient(low = "blue", high = "red") +
  labs(title = "Car Traffic Volume Over Time Trend",
       x = "Days Since Start",
       y = "Daily Car Traffic",
       color = "Daily Traffic") +
  theme_minimal()

@@

The scatter plot illustrates the fluctuating daily car counts over time, with a trend line suggesting an initial increase in traffic followed by a decline.

@@

This study uses negative binomial regression to address the problem of over-dispersion, which occurs when observed variance is greater than the mean and is frequently seen in count data. The dependent variable in the model is the daily count of cars, which is assumed to be impacted by temporal variations and the counts of alternative transportation modes (such as buses, bicycles, and pedestrians).

@@

There are two reasons for using negative binomial regression. First, since events tend to cluster together in both place and time, traffic count data usually show overdispersion. For such data, the negative binomial model makes sense since it allows the variance to surpass the mean. Second, given that traffic counts frequently deviate from a normal distribution, this model can account for their skewed nature.

The analysis is divided into two temporal models in order to evaluate the effects of the Route 508 Lake Shore streetcar service upgrades: `model_before`, which is the model before the improvements, and `model_after`, which is the model after the improvements. This method clarifies the relationship that existed before and after the policy intervention between the independent and dependent variables.

Together with a temporal trend, the `model_before` captures the baseline relationship between automobile traffic and other modes of transportation. As an illustration of possible changes in traffic patterns brought about by the improved streetcar service, the `model_after`, on the other hand, depicts the altered dynamics following the enhancement. By use of comparative analysis, these models demonstrate the effectiveness of the Route 508 enhancements in mitigating car congestion.

@@

```{r residuals-before-model, fig.cap="Residuals of the negative binomial regression model before and after the transit improvement. This plot helps to visualize if there are any patterns or systematic deviations unexplained by the model.", echo=FALSE, message=FALSE, warning=FALSE, fig.align='center'}
# Read and prepare data
analysis_data <- read.csv(here("data", "analysis_data", "cleaned-data.csv"))

@@

# Visualizing residuals to identify patterns
par(mfrow = c(1, 2))
plot(residuals(model_before), type = "p", main = "Residuals of Model Before Improvement")
plot(residuals(model_after), type = "p", main = "Residuals of Model After Improvement")

# Examining the normality of residuals
# hist(resid(model_before), main = "Histogram of Residuals - Before")
# hist(resid(model_after), main = "Histogram of Residuals - After")
```

Figures 6 and 7 show an examination of residuals from the negative binomial regression models before and after the public transit improvement. For the model before improvements, the residuals plot does not indicate any apparent patterns or systemic deviations, suggesting that the model's assumptions may be adequate for the data. After the transit improvements, the residuals' plot again shows no obvious patterns, which suggests a random scatter.

```{r model summaries, echo=FALSE, message=FALSE, warning=FALSE}

# Read and prepare data
analysis_data <- read.csv(here("data", "analysis_data", "cleaned-data.csv"))
analysis_data$count_date <- as.Date(analysis_data$count_date)

@@

analysis_data <- analysis_data %>%
  mutate(time_trend = as.numeric(count_date - min(count_date, na.rm = TRUE)))

# Building negative binomial regression models for before and after improvement
model_before <- glm.nb(daily_cars ~ daily_bus + daily_peds + daily_bike + time_trend, data = before_improvement)
model_after <- glm.nb(daily_cars ~ daily_bus + daily_peds + daily_bike + time_trend, data = after_improvement)

# Tidying the model summaries 
tidy_model_before <- tidy(model_before)
tidy_model_after <- tidy(model_after)

# Model before improvement
kable(tidy_model_before, format = "latex", booktabs = TRUE, caption = "Negative Binomial Regression Model - Before Improvement") %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = F) %>%
  column_spec(1, bold = TRUE, width = "3cm") %>%
  column_spec(2, bold = TRUE, color = "blue", width = "3cm") %>%
  column_spec(3, width = "3cm")

kable(tidy_model_after, format = "latex", booktabs = TRUE, caption = "Negative Binomial Regression Model - After Improvement") %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = F) %>%
  column_spec(1, bold = TRUE, width = "3cm") %>%
  column_spec(2, bold = TRUE, color = "blue", width = "3cm") %>%
  column_spec(3, width = "3cm")

@@

Table 2 represents the traffic patterns before the improvement. These results produced significant coefficients for both the intercept and daily bus traffic, as well as the time trend variable. The intercept, representing the baseline log-count of cars when all other variables are zero, was significant with an estimate of 9.340 and a standard error of 0.072, indicating a high baseline traffic volume. Daily bus traffic had a positive and significant relationship with car traffic, with a coefficient of 0.261, suggesting that increases in bus traffic were associated with increases in car traffic. The time trend also showed a significant, albeit small, positive effect on car traffic `(β = 0.000287)`, indicating a gradual increase in car traffic over time.

Pedestrian and bicycle traffic, however, were not significantly associated with car traffic before the improvements (p \> 0.05 for both).

Post-improvement, the model exhibits significant changes in the dynamics of traffic patterns. The intercept remains significant but increases to 12.283, reflecting an overall higher baseline of car traffic possibly due to other factors. The coefficient for daily bus traffic notably decreases to 0.096, indicating a diminished relationship between bus and car traffic after the public transit improvements.

The time trend shifts from a slight increase to a significant decrease `(β = -0.001033)`, reflecting a reversal in the car traffic trend, which now shows a significant decline over time post-improvement.

Variables for pedestrian and bicycle traffic remained non-significant, suggesting that these modes of transport did not have a substantial impact on car traffic volumes in either model.

The model results indicate a significant reduction in the relationship between bus traffic and car congestion following the transit improvements, as shown by the decrease in the coefficient of daily bus traffic from 0.261 to 0.096. This suggests that improvements in the public transit system might be effectively alleviating car congestion. Additionally, the significant negative time trend in the post-improvement model underscores a successful overall reduction in car traffic between 2012 to 2020.

@@

These local findings align with research conducted by Verbavatz and Barthelemy `source` in which they found that urban area size and public transport density are pivotal in managing car traffic in cities, rather than urban density alone. The article further states that enhancing the density and accessibility of public transport is a viable path to mitigating traffic and its associated environmental impacts.

Moreover, the benefits of improved public transit does not stop at decreased car congestion. Insights from the American Public Transportation Association, which illustrate that public transit investments yield substantial congestion relief and economic benefits. Specifically, every dollar spent on public transportation generates 5 dollars in economic return and an associated creation of approximately 50,000 jobs for every billion dollars invested `source`. While on the individual level, households in America can save more than \$13,000 per year by using public transit and owning one less car.

These advantages of public transportation investments are in alignment with the principles of sustainable development as defined by the UN's World Commission on Environment and Development: to meet current needs without compromising the ability of future generations to meet their own needs `source`. Together, these strands of evidence support the case for enhancing public transit as a strategic approach to improving urban traffic conditions and achieving broader sustainability goals.

@@

Supporting this claim, Lexer Quamie's research (2011) `source` highlights the integral role of transportation policy in civil rights, particularly in periods of pronounced unemployment and income disparity. Quamie emphasizes that transportation decisions significantly affect who can access employment opportunities and essential services, positioning public transit as a pivotal tool in the fight for civil rights. By facilitating or hindering access to critical resources, transportation policies not only impact economic outcomes but also health and environmental conditions. This is evident as inadequate or unaffordable public transit disproportionately affects low-income populations, limiting their ability to reach jobs, healthcare, and thereby affecting their overall economic and health status. Additionally, prioritization of car travel over public transit can lead to increased emissions, affecting urban air quality and contributing to environmental degradation.

@@

# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check

## Diagnostics

>>>>

---
title: "The Route 508 Streetcar and Congestion on Toronto's Shoreline, 2011 - 2020"
subtitle: "Examining Traffic Patterns Before and After Public Transit Access was Reinstated"
author: 
  - Maria Mangru
thanks: "Code and data are available at: https://github.com/MariaMangru/Traffic-Congestion-on-Toronto-s-Shoreline."
date: today
date-format: long
abstract: "This study investigates the impact of access to the Route 508 Lake Shore streetcar in Toronto on reducing urban car congestion from 2011 to 2020. By analyzing traffic volume data from the City of Toronto's Transportation Services Division and employing negative binomial regression models, we observed a notable shift in travel behavior, highlighted by a significant decrease in the correlation between bus traffic and car congestion. Specifically, the coefficient for daily bus traffic reduced from 0.261 to 0.096 after access was reinstated, indicating a reduction in car traffic as more public transit options became available. This highlights the effectiveness of enhancing public transit infrastructure not only in easing car congestion but also in contributing to broader environmental and economic benefits. By improving public transit, cities can potentially achieve sustainable development goals by providing equitable access to essential services and reducing environmental impacts, thereby improving quality of life for all urban residents."
format: 
  pdf:
    documentclass: article
    geometry: [margin=1in]
number-sections: true
bibliography: references.bib
---

```{r, include= FALSE, echo=FALSE, message=FALSE}
#| include: false
#| warning: false
#| message: false

library(dplyr)
library(ggplot2)
library(tidyr)
library(readr)
library(here)
library(knitr)
library(kableExtra)
library(stringr)
library(broom)
library(MASS) 
library(car)
library(patchwork)
```

```{r, include= FALSE, echo=FALSE, message=FALSE}

#Import cleaned data 
analysis_data <- read_csv(here("data", "analysis_data", "cleaned-data.csv"))

@@

The primary aim of this research is to evaluate the effect of the 508 Lake Shore streetcar enhancements on daily traffic patterns, focusing on how these changes influence the interaction between car traffic and other transportation modes such as buses, bicycles, and pedestrians. By employing traffic volume data collected by the City of Toronto's Transportation Services Division, we utilize negative binomial regression models to quantify the shifts in traffic patterns from 2011 to 2020. The estimand of this study is the change in the correlation between daily bus traffic and daily car traffic following the transit improvements. For the purposes of this research, the terms "improvements" and "reinstating of the streetcar" are used interchangeably. Both terms refer to the suite of upgrades and changes implemented along the 508 Lake Shore route, which included the restoration of streetcar services that had previously been discontinued or altered.'

@@

This research relies on traffic volume provided by the City of Toronto's Transportation Services Division, available through the Toronto OpenData portal. The dataset, titled "Traffic Volumes at Intersections for All Modes," is an open-access resource under the Open Government License - Toronto, ensuring broad usability and transparency. The data files for the years 2010-2019 and 2020-2029 were utilized, however, the data files are uploaded daily with data ranging from 1980 to present. The traffic volume data are cross-referenced with the City of Toronto's Street Centreline, Intersection File, and Street Traffic Signal datasets.

@@

The analysis in this paper was conducted using the R programming language @citeR, which facilitated the statistical computations and data visualization. Further, a suite of R packages, including several from the tidyverse collection were utilized. The @dplyr package was used the transformation and summarization of traffic data. For importing and processing string data, @readr and @stringr were utilized. The handling of date and time variables was done using @lubridate and data visualization was achieved with @ggplot2, allowing us to explore and reveal traffic patterns visually. Additionally, @tidyverse assisted in reshaping the data to ensure it was structured appropriately for analysis. File path management was handled using the @here package, which was essential for ensuring reproducibility across different computing environments.

The dataset was refined to aggregate daily totals for each mode of transportation, with the data aligned and merged based on date, location, and traffic mode specifics. While the dataset was merged from two data files which span from 2010 to 2019 and 2020 to 2029, this research will focus on December 2011 to December 2014 and November 2018 to February 2020. These years were chosen to provide a comprehensive view of traffic patterns prior to, during, and following the enhancement of the 508 Lake Shore route. The data collection period was intentionally concluded in February 2020 to ensure that the results were not influenced by the disruptions caused by the COVID-19 pandemic.

@@

In this study, the examination of traffic dynamics is conducted by analyzing interactions among various modes of transportation at various intersections. As such, traffic counts for each mode of transit were aggregated to daily totals to allow for better analysis. For example, car traffic variables such as `sb_cars_l`, `nb_cars_r`, `eb_cars_t`, among others, representing different directions and types of car movements at intersections, were combined to form the Daily Car Traffic (`daily_cars`). This variable captures the total volume of car traffic and serves as a primary indicator of vehicular congestion. Similarly, this was done to create the variable Daily Bus Traffic (`daily_bus`), which helps in evaluating the volume and effectiveness of bus transit in relation to car traffic. Furthermore, Daily Pedestrian Traffic (`daily_peds`) and Daily Bicycle Traffic (`daily_bike`) were created to aid in understanding the prevalence and trends of non-vehicular traffic modes. A Time Trend variable (`time_trend`) was constructed to analyze the longitudinal changes in traffic patterns over time, providing insights into the temporal effects of transit improvements.

```{r summary-stats-table, echo=FALSE, message=FALSE, warning=FALSE}
#| label: tbl-trafficstats
#| tbl-cap: Summary statistics of Daily Traffic Volume by mode of transportation. This table displays the mean and median counts of cars, buses, pedestrians, and bicycles

# Generate summary statistics for traffic data
summary_stats <- summarise(analysis_data,
  Mean_Cars = mean(daily_cars, na.rm = TRUE),
  Median_Cars = median(daily_cars, na.rm = TRUE),
  Mean_Buses = mean(daily_bus, na.rm = TRUE),
  Median_Buses = median(daily_bus, na.rm = TRUE),
  Mean_Pedestrians = mean(daily_peds, na.rm = TRUE),
  Median_Pedestrians = median(daily_peds, na.rm = TRUE),
  Mean_Bicycles = mean(daily_bike, na.rm = TRUE),
  Median_Bicycles = median(daily_bike, na.rm = TRUE)
)

@@

The summary statistics presented in @tbl-trafficstats offers a quantitative comparison across different modes of transportation, capturing the daily traffic volumes on the 508 Lake Shore route. Notably, **Daily Car Traffic** significantly surpasses other modes of transit, with mean and median values at 14,169.85 and 13,291.5, respectively. This highlights the predominant use of cars in the area, highlighting the challenges of reducing vehicular congestion.

In contrast, **Daily Bus Traffic** shows much lower volumes, with a mean of 223.91 and a median of 190.5, reflecting less frequent usage but indicating a critical focus area for transit improvements. **Daily Pedestrian Traffic** and **Daily Bicycle Traffic** also exhibit considerably lower traffic volumes, with means of 686.06 and 50.49, respectively. The pedestrian traffic presents a median significantly lower than its mean, at 344, suggesting a skewed distribution with potentially high traffic on certain days or specific intersections. Similarly, the median for bicycle traffic stands at 22, indicating that on many days, bicycle usage is quite low despite a higher average, which might reflect seasonal variations or special events influencing bike use.

```{r}
#| label: fig-trafficplots
#| fig-cap: Traffic Trends by Mode along the 508 Lakeshore Route from December 2011 to December 2014 and November 2018 to February 2020.
#| echo: FALSE
#| warning: FALSE
#| message: FALSE

# Base theme for all plots 
base_theme <- theme_minimal() +
  theme(plot.title = element_text(size = 10), 
        axis.title = element_text(size = 8), 
        axis.text = element_text(size = 6))

# Car Traffic Plot
line_graph_car <- analysis_data %>%
  ggplot(aes(x = count_date, y = daily_cars)) +
  geom_smooth(se = FALSE) +  
  labs(title = "Car Traffic",
       x = "Date",
       y = "Total Cars") +
  base_theme

# Bus Traffic Plot
line_graph_buses <- analysis_data %>%
  ggplot(aes(x = count_date, y = daily_bus)) +
  geom_smooth(se = FALSE) +
  labs(title = "Bus Traffic",
       x = "Date",
       y = "Total Buses") +
  base_theme

# Pedestrian Traffic Plot
line_graph_pedestrians <- analysis_data %>%
  ggplot(aes(x = count_date, y = daily_peds)) +
  geom_smooth(se = FALSE) +
  labs(title = "Pedestrian Traffic",
       x = "Date",
       y = "Total Pedestrians") +
  base_theme

# Cyclist Traffic Plot
line_graph_cyclists <- analysis_data %>%
  ggplot(aes(x = count_date, y = daily_bike)) +
  geom_smooth(se = FALSE) +
  labs(title = "Cyclist Traffic",
       x = "Date",
       y = "Total Cyclists") +
  base_theme

# Combine plots 
combined_plots <- (line_graph_car | line_graph_buses) / 
                  (line_graph_pedestrians | line_graph_cyclists) & 
                  plot_layout(guides = 'collect')

combined_plots

@@

@fig-trafficplots presents a visual analysis of the various modes of traffic from 2011 to 2020, demonstrating their respective trends and allowing for comparative insights. The car traffic graph illustrates an initial uptrend in volume, reaching a peak around 2016, followed by a pronounced decline that coincides with the reintroduction of the Route 508 Lake Shore streetcar, suggesting a potential correlation with improved public transit and a reduction in car use. Concurrently, bus traffic mirrors this pattern to a degree, with a rise and subsequent fall post-2016, hinting at an adaptive response in public transport usage. The pedestrian traffic trendline depicts a mid-period dip followed by a recovery and then a decline again. Cyclist traffic, on the other hand, displays a waveform pattern with significant ups and downs, possibly reflecting the impact of varied factors including public attitudes towards cycling, the introduction or enhancement of cycling infrastructure, and seasonal changes that affect cycling prevalence.

```{r}
#| label: fig-cartimegraph
#| fig-cap: Daily Car Traffic Over Time. This graph displays the general trend in number of cars from the beginning of data collection period.
#| echo: FALSE
#| warning: FALSE
#| message: FALSE

# Visualizing car traffic with respect to the time trend
time_trend_graph_car <- analysis_data %>%
  ggplot(aes(x = time_trend, y = daily_cars)) +
  geom_point(aes(color = daily_cars), alpha = 0.5) + 
  geom_smooth(method = "lm", se = FALSE, color = "blue") +  
  scale_color_gradient(low = "blue", high = "red") +
  labs(title = "Car Traffic Volume Over Time Trend",
       x = "Days Since Start",
       y = "Daily Car Traffic",
       color = "Daily Traffic") +
  theme_minimal()

@@

The scatter plot in @fig-cartimegraph, representing daily car traffic counts over time against a time trend, offers a granular view of the data. Each dot signifies the traffic volume on a single day, while the trend line captures the general trajectory over the entire data collection period. An initial increase in traffic volume is apparent, eventually giving way to a downward trend.

@@

Modeling traffic dynamics poses unique challenges, particularly with count data that often exhibit over-dispersion—where the variance exceeds the mean—making conventional regression models like Poisson unsuitable. To address this, this study employs a negative binomial regression model in order to handle over-dispersion and more aligned with the distribution characteristics of traffic counts.

We define our model with the daily car traffic count as the dependent variable, hypothesizing its dependence on several factors: temporal trends and daily counts of alternative transport modes—buses, pedestrians, and bicycles.

@@

The rationale for selecting the negative binomial model is based on two primary areas. Firstly, traffic data are prone to clustering in both spatial and temporal dimensions, leading to overdispersion. Secondly, traffic count data typically skew away from a normal distribution, and the negative binomial model is equipped to address this skewness.

To capture the effects of the Route 508 Lake Shore streetcar service reinstatement, the analysis dives into two temporal segments: pre-improvement (`model_before`) and post-improvement (`model_after`). This segmentation enables a nuanced understanding of the relationships between car traffic and alternative transport modes before and after the streetcar was reinstated. The `model_before` sets the baseline, showcasing the original dynamics between cars and other transport forms. The `model_after` reveals the traffic landscape streetcar service has resumed. This offers a comparative view to measure the efficacy of these public transit access improvements in mitigating car congestion.

This modeling approach not only provides insights into the immediate effects of transit policy changes but also establishes a methodological framework for future studies examining the longitudinal impact of urban transit interventions.

@@

```{r}
#| label: fig-residualsmodel
#| fig-cap: Residuals of the negative binomial regression model before and after the transit improvement. This plot helps to visualize if there are any patterns or systematic deviations unexplained by the model.
#| echo: FALSE
#| warning: FALSE
#| message: FALSE

@@

# Visualizing residuals to identify patterns
#par(mfrow = c(1, 2))
#plot(residuals(model_before), type = "p", main = "Residuals of Model Before Improvement")
#plot(residuals(model_after), type = "p", main = "Residuals of Model After Improvement")

# Adjust overall plot margins (bottom, left, top, right) and spacing for multi-figure plots
par(mfrow = c(1, 2), mar = c(5, 4, 4.5, 2) + 0.1, mgp = c(2, 0.7, 0), tcl = -0.3, cex.main = 0.9)

# Visualizing residuals for the 'before improvement' model
plot(residuals(model_before), 
     type = "p", 
     main = "Residuals Before Improvement", 
     xlab = "Index", 
     ylab = "Residuals", 
     xlim = c(0, max(length(residuals(model_before)), length(residuals(model_after)))),
     ylim = range(c(residuals(model_before), residuals(model_after))))

# Visualizing residuals for the 'after improvement' model
plot(residuals(model_after), 
     type = "p", 
     main = "Residuals After Improvement", 
     xlab = "Index", 
     ylab = "Residuals", 
     xlim = c(0, max(length(residuals(model_before)), length(residuals(model_after)))),
     ylim = range(c(residuals(model_before), residuals(model_after))))

The residual plots depicted in @fig-residualsmodel serve as a diagnostic tool to evaluate the adequacy of the negative binomial regression models used in this study, both before and after the implementation of public transit improvements. The left panel of the figure illustrates the residuals for the model before the reinstatement of the Route 508 Lake Shore streetcar, showing a random dispersion of data points around the zero line. This lack of systematic pattern suggests that the model assumptions are met, indicating a good fit for the pre-improvement traffic data.

Similarly, the right panel of the figure, representing the residuals post-improvement, exhibits a scatter of data points without discernible trends or biases. The randomness observed in the distribution of these residuals supports the conclusion that the post-improvement model is also well-fitted to the data, capturing the essential variability in daily car counts without leaving unexplained structure in the residuals.

```{r}
#| echo: FALSE
#| warning: FALSE
#| message: FALSE

# Read and prepare data
analysis_data$count_date <- as.Date(analysis_data$count_date)

@@

analysis_data <- analysis_data %>%
  mutate(time_trend = as.numeric(count_date - min(count_date, na.rm = TRUE)))
```

```{r}
#| label: tbl-modelbeforesummaries
#| tbl-cap: "Model Summary of Traffic Patterns Before Improvement on the 508 Lakeshore Route"
#| echo: FALSE
#| warning: FALSE
#| message: FALSE

# Tidying the model summaries 
tidy_model_before <- broom::tidy(model_before)

# Model before improvement table
kable(tidy_model_before, format = "latex", booktabs = TRUE) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
  column_spec(1, bold = TRUE, width = "3cm") %>%
  column_spec(2, color = "blue", width = "3cm") %>%
  column_spec(3, width = "3cm") %>%
  column_spec(4, width = "3cm") %>%
  column_spec(5, width = "3cm")

@@

The pre-improvement model, as detailed in @tbl-modelbeforesummaries, reveals significant coefficients for both the intercept and daily bus traffic, alongside a temporal trend. With a coefficient of 9.34, the model showcases a high baseline volume of car traffic, independent of other variables. Bus traffic, represented by a coefficient of 0.261, appears to move in tandem with car traffic, implying that increases in bus usage were historically linked to increases in car traffic. However, the coefficients for pedestrian and bicycle traffic do not present a significant association with car traffic, reflecting a p-value greater than 0.05 for both.

```{r}
#| label: tbl-modelaftersummaries
#| tbl-cap: "Model Summary of Traffic Patterns After Improvement on the 508 Lakeshore Route"
#| echo: FALSE
#| warning: FALSE
#| message: FALSE

# Building negative binomial regression models for before and after improvement
model_after <- glm.nb(daily_cars ~ daily_bus + daily_peds + daily_bike + time_trend, data = after_improvement)

# Tidying the model summaries 
tidy_model_after <- broom::tidy(model_after)

# Model after improvement table
kable(tidy_model_after, format = "latex", booktabs = TRUE) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
  column_spec(1, bold = TRUE, width = "3cm") %>%
  column_spec(2, color = "blue", width = "3cm") %>%
  column_spec(3, width = "3cm") %>%
  column_spec(4, width = "3cm") %>%
  column_spec(5, width = "3cm")

Conversely, the post-improvement model shown in @tbl-modelaftersummaries exhibits a marked change in these dynamics. The increase in the intercept to 12.283 suggests an elevated baseline of car traffic, potentially as a result of external influences not accounted for in the model. The coefficient for daily bus traffic experiences a considerable reduction to 0.096, signaling a weakened relationship between bus and car traffic in the wake of the transit improvements. This could indicate a disconnect of these modes of transport, possibly due to more passengers opting for the improved public transit system over personal vehicles. The time trend coefficient inverses to -0.001033, denoting a statistically significant decline in car traffic over time post-improvement, showcasing the potential efficacy of the public transit enhancements in reducing car traffic.

These findings show that following the public transit improvements, the relationship between bus traffic and car congestion significantly diminished, potentially reflective of the transit system's capacity to alleviate car congestion. Additionally, the persistent non-significance of pedestrian and bicycle traffic variables implies these modes have not impacted car traffic volumes in the models considered.

@@

These local findings align with research conducted by @Verbavatz2019 in which they found that urban area size and public transport density are pivotal in managing car traffic in cities, rather than urban density alone. The article further states that enhancing the density and accessibility of public transport is a viable path to mitigating traffic and its associated environmental impacts.

Moreover, the benefits of improved public transit does not stop at decreased car congestion. Insights from the @APTA2023, which illustrate that public transit investments yield substantial congestion relief and economic benefits. Specifically, every dollar spent on public transportation generates 5 dollars in economic return and an associated creation of approximately 50,000 jobs for every billion dollars invested . While on the individual level, households in America can save more than \$13,000 per year by using public transit and owning one less car.

These advantages of public transportation investments are in alignment with the principles of sustainable development as defined by the UN's World Commission on Environment and Development: to meet current needs without compromising the ability of future generations to meet their own needs (@Lachman2013Chapter3). Together, these strands of evidence support the case for enhancing public transit as a strategic approach to improving urban traffic conditions and achieving broader sustainability goals.

@@

Supporting this claim, @Quamie2011 research highlights the integral role of transportation policy in civil rights, particularly in periods of pronounced unemployment and income disparity. Quamie emphasizes that transportation decisions significantly affect who can access employment opportunities and essential services, positioning public transit as a pivotal tool in the fight for civil rights. By facilitating or hindering access to critical resources, transportation policies not only impact economic outcomes but also health and environmental conditions. This is evident as inadequate or unaffordable public transit disproportionately affects low-income populations, limiting their ability to reach jobs, healthcare, and thereby affecting their overall economic and health status. Additionally, prioritization of car travel over public transit can lead to increased emissions, affecting urban air quality and contributing to environmental degradation.

@@

## Motivation

For what purpose was the dataset created? Was there a specific task in mind? Was there a specific gap that needed to be filled? Please provide a description.
